Concurrent online sampling, for all, for free

### 2022-07-17 paper/main.tex
[wip] paper

        % How does the head pointer look like (LOOOK A THIS)
        The free list and the LoS can be accessed using a head pointer. A pointer to a node is an index into the preallocated node are. Threads can only push a node or pop a node from a list. To ensure that no concurrent modifications can be meade to a list the head pointer is written with one atomic compare and swap operation. In order to detected changes made by other threads, a thread first needs to store the current pointer to a list. When the pointer changed in the meantime the compare and swap operation will fail. However, nodes are frequently added and removed to a list and it could be the case that the pointer was changed and then changed back to the original node. In this case the compare and swap operation would fail but the pointer was modified in the meantime. To address this problem the pointer also contains a version besides the index. This pointer can be stored in 64 bit integer, hereby the first 32bit store the version and the rest stores the index. Figure \ref{fig:pointer} visualizes the layout of a pointer.
        \begin{figure}[h]
            \includegraphics[height=2.2cm]{figure4.pdf}
            \caption{The structure of a head pointer}
            \label{fig:pointer}
        \end{figure}

        % Actual acquiring
        The process of acquire a node from the LoS can be broken down into three steps. First the thread needs to read the head pointer and store the current value. Afterwards, the thread can use the index stored in the pointer to retrieve the node form the LoS. In the second step, the thread needs to generate the successor for the Skip stored in the node and lastly the thread tries to writes the successor into the head of the LoS. If the update succeeds the thread now owns a new Skip and the head of the LoS is set to the successor Skip. On the other hand, if the update fails, the thread needs to start over. However, in practice this is not a problem, because every time a thread fails to update the head, another thread successfully updated the head and therefore overall progress is guaranteed. Figure \ref{fig:acquire} shows the three steps necessary to acquire a Skip and the retry loop.
        \begin{figure}[h]
            \includegraphics[height=2.4cm]{figure5.pdf}
            \caption{Flow diagram that shows the process of acquire a Skip from the LoS}
            \label{fig:acquire}
        \end{figure}

        % Generating the successor
        If the LoS only contains one node, the thread needs to generate a successor before takeing ownership of this node. In order to generate a Skip the thread needs to own a node, because nodes can only be modified when they are owned by a thread. It could be the case that the thread already owns a node from a previous iteration. On the other hand, if the thread does not own a node it needs to retrieve one from the free list. Using a node and the current Skip at the head of the LoS the thread can generate the successor. However, it could also be the case that the LoS contains more then one node and no successor needs to be generated. But if the thread already owns a node, it needs to deallocate this node. Because a thread cannot own more then one node. Deallocateing a node refers to inserting the node into the free list. To summarize, there a two cases where a thread needs to allocate or deallocate a node: 
        \begin{itemize}
            \item Allocate: If the thread does not own a node and the LoS only contains one node
            \item Deallocate: If the thread owns a node and the LoS contains more the one node
        \end{itemize}

@@

    \section{Example}
        % Introduction to the example 
        The following example will illustrate hwo the algorithm could work with two inserting threads, $T0$ and $T1$. Figure \ref{fig:example} visualizes the state of all nodes and Skips used in this example. This example assumes that the sample as already been prefilled.
        \begin{figure}[h]
            \includegraphics[height=8cm]{figure6.pdf}
            \caption{The state of every Skip and node for the example}
            \label{fig:example}
        \end{figure}

@@

        This example contains both allocation and deallocation. $T0$ needs to allocate a node in step 1 and $T1$ needs to allocate a node in step 2. Because $T1$ returns its Skip, the LoS contains two Skips after Step 4. Therefore, $T0$ needs to deallocate its own node in Step 5.

@@

        % By design / const memory
        (1) By design the algorithm takes every inserted tuple into consideration and decides whether the tuple is a sample tuple or not. Therefore, the sample is always updated. (3) Also, because all nodes are preallocated no more memory allocations are required after the algorithm is initialized and the amount of memory only scales with the maximum amount of inserting threads. Thus the memory usage is constant throughout the runtime of the algorithm.

        % Single compare and swap operation
        (5) The algorithm reduces the amount of extra synchronization to a minimum. There are only tow cases where synchronization is required and all of the synchronization can be done with simple compare and swap (CAS) operations. CAS operations are a very efficient way to  synchronization multiple threads and most of the time a single CAS operation is enough to acquire a new Skip. Because it is likely that a thread already owns a node from the previous iteration and no additional node needs to be retrieved from the free list. In practice a CAS operation could fail if multiple threads access the same list. However, if one CAS operation fails the next one is more likely to succeed, because the successful thread needs to process the Skip first before it access the data structure again. The other case that requires synchronization is, when writing to the sample. But because the synchronization is very fine grained most of the time threads do not need to wait to acquire the write lock for the sample.

        % Cheap operations
        (2) Most of the additional overhead introduced by the algorithm, is acquiring a Skip. When a thread already owns a Skip, only one additional decrement and compare operation needs to be execute. Because most of the time a thread will own a Skip, the introduced overhead is very low. This sampling algorithm was also integrated into the DBMS Umbra\cite{Umbra} and figure \ref{fig:performance} visualizes the throughout with and without sampling. Most of the time the performance is exactly the same, only in some cases the throughout with sampling off is slightly higher. This confirms the claim that online sampling introduced virtually no overhead.
        \begin{figure}[h]
            \includegraphics[height=4.6cm]{figure7.pdf}
            \caption{TPC-H bulk-loading throughput in Umbra\cite{Umbra} with
            and without sampling \cite[figure 6]{OG}}
            \label{fig:performance}
        \end{figure}

        % Simple interface
        (4) Integrating this sampling algorithm into an existing DBMS can be done with a simple interface. To initialize the algorithm only the node array needs to be allocated and the pointer the free list and the LoS needs to be created. Acquiring a Skip can be done with one simple function call that handles everything, like accessing the LoS and allocating or deallocating nodes. Also, deciding whether a tuple is a sample tuple or not and inserting this tuple into the sample can be done with one simple function call. In total the interface of the algorithm consist only of a few simple function calls that need be from the existing code.

>>>>

        % How does the head pointer look like
        The free list and the LoS can be accessed using a pointer to the first node in the linked list, also called a head pointer. A pointer to a node is an index into the preallocated node array. Additionally, a node can only be modified when it is owned by a thread. To ensure that no concurrent modifications can be meade to a list the head pointer is written with one atomic compare and swap (CAS) operation. 
        However, if a thread reads from a node stored in the LoS and wants to make changes to the list based on the data. Another thread could make changes to the list in the meantime. In order to detected these changes made by other threads, a thread first needs to store the current head pointer for a list. When the pointer changed in the meantime the CAS operation will fail. However, nodes are frequently added and removed to a list and it could be the case that the head pointer was changed and then changed back to the original node. In this case the CAS operation would not fail, but the head pointer was modified in the meantime. To address this problem the head pointer for a list also contains a version besides the index. A head pointer can be stored in 64 bit integer, hereby the first 32 bit store the version and the rest stores the index. Figure \ref{fig:pointer} visualizes the layout of a pointer.
        \begin{figure}[h]
            \includegraphics[height=2.2cm]{figure4.pdf}
            \caption{The structure of a head pointer}
            \label{fig:pointer}
        \end{figure}

        % Actual acquiring
        The process of acquire a node from the LoS can be broken down into three steps. First the thread needs to read the head pointer and store the current value. Afterwards, the thread can use the index stored in the pointer to retrieve the node form the LoS. In the second step, the thread needs to generate the successor for the Skip stored in the node.Lastly, the thread tries to set the successor as the head of the LoS. If the update succeeds the thread now owns a new Skip and the head of the LoS is set to the successor Skip. On the other hand, if the update fails, the thread needs to start over. However, in practice this is not a problem, because every time a thread fails to update the head, another thread has successfully updated the head and therefore overall progress is guaranteed. Figure \ref{fig:acquire} shows the three steps necessary to acquire a Skip and the retry loop.
        \begin{figure}[h]
            \includegraphics[height=2.4cm]{figure5.pdf}
            \caption{Flow diagram that shows the process of acquiring a Skip from the LoS}
            \label{fig:acquire}
        \end{figure}

        % Generating the successor
        If the LoS only contains one node, the thread needs to generate a successor before takeing ownership of the node. In order to generate a Skip the thread needs to own a node, because nodes can only be modified when they are owned by a thread. It could be the case that the thread already owns a node from a previous iteration. On the other hand, if the thread does not own a node it needs to retrieve one from the free list. Using a node and the current Skip at the head of the LoS the thread can generate the successor. However, it could also be the case that the LoS contains more then one node and no successor needs to be generated. But if the thread already owns a node, it needs to deallocate this node. Because a thread cannot own more then one node. Deallocating a node refers to inserting the node into the free list. To summarize, there a two cases where a thread needs to allocate or deallocate a node: 
        \begin{itemize}
            \item Allocate: If the thread does not own a node and the LoS only contains one node
            \item Deallocate: If the thread owns a node and the LoS contains more the one node
        \end{itemize}

@@

    \subsection{Preloading}
        % Recap
        During the sampling phase the algorithm assumes that the sample is full. Initially the sample is empty and therefore the sample needs to be prelaoded. To preload a sample of size $m$ the first $m$ tuples which are inserted into the database are also inserted into the sample. The sample is preloaded in the second phase of the algorithm before the actual sampling phase starts.

        % How does preloading work
        The size of the sample $m$ is fixed. Thus a counter $v$ can be used to decide whether a tuple needs to be inserted into the sample during the preload phase. This counter is initially set to zero. When a thread receives a tuple it first checks whether $v<m$. If $v>=m$ the preload phase is over and the thread continues with the normal sampling procedure described above. On the other hand, if $v<m$ the thread performance an atomic fetch an increment operation on the counter. Afterwards, the fetched value $v'$ could be greater or equal to $m$ because another thread could have incremented the counter in the meantime. However, if $v'<m$ the tuple should be inserted at index $v'$ into the sample. The tuple is only written to the sample if the row at index $v'$ is still empty. Because another thread could have written into this row as part of the normal sample procedure if the preloading phase ended. After the prelaod phase is over a thread does not have to check to counter again.

    \section{Example}
        % Introduction to the example 
        The following example will illustrate how the algorithm could work with two inserting threads, $T0$ and $T1$. Figure \ref{fig:example} visualizes the state of all nodes and Skips used in this example. This example assumes that the sample as already been preloaded.
        \begin{figure}[h]
            \includegraphics[height=8cm]{figure6.pdf}
            \caption{The state of every Skip and node for the example}
            \label{fig:example}
        \end{figure}

@@

        This example contains both allocation and deallocation of a node. $T0$ needs to allocate a node in step 1 and $T1$ needs to allocate a node in step 2. Because $T1$ returns its Skip, the LoS contains two Skips after Step 4. Therefore, $T0$ needs to deallocate its own node in Step 5.

@@

        % By design / const memory
        (1) By design the algorithm takes every inserted tuple into consideration and decides whether the tuple is a sample tuple or not. Therefore, the sample is always updated. (3) Also, because all nodes are preallocated no more memory allocations are required after the algorithm is initialized. The amount of memory only scales with the maximum amount of inserting threads. Thus, the memory usage is constant throughout the runtime of the algorithm.

        % Single compare and swap operation
        (5) The algorithm reduces the amount of extra synchronization to a minimum. There are only tow cases where synchronization is required and all of the synchronization can be done with simple compare and swap (CAS) operations. CAS operations are a very efficient way to  synchronization multiple threads and most of the time a single CAS operation is enough to acquire a new Skip. Because it is likely that a thread already owns a node from the previous iteration and therefore no additional node needs to be retrieved from the free list. In practice a CAS operation could fail if multiple threads access the same list. However, if one CAS operation fails the next one is more likely to succeed, because the successful thread needs to process the Skip first before it access the data structure again. The other case that requires synchronization is, when writing to the sample. But because the synchronization is very fine grained most of the time threads do not need to wait to acquire the write lock for a sample entry.

        % Cheap operations
        (2) Most of the additional overhead introduced by the algorithm, is acquiring a Skip. When a thread already owns a Skip, only one additional decrement and compare operation needs to be execute. Because most of the time a thread will own a Skip, the introduced overhead is very low. This sampling algorithm was also integrated into the DBMS Umbra\cite{Umbra} and figure \ref{fig:performance} visualizes the throughput with and without sampling. Most of the time the performance is exactly the same, only in some cases the throughput with sampling off is slightly higher. This confirms the claim that online sampling introduced virtually no overhead.
        \begin{figure}[h]
            \includegraphics[height=4.6cm]{figure7.pdf}
            \caption{TPC-H bulk-loading throughput in Umbra\cite{Umbra} with
            and without sampling \cite[figure 6]{OG}}
            \label{fig:performance}
        \end{figure}

        % Simple interface
        (4) Integrating this sampling algorithm into an existing DBMS can be done with a simple interface. To initialize the algorithm only the node array needs to be allocated and the pointer the free list and the LoS need to be created. Acquiring a Skip can be done with one simple function call that handles everything, like accessing the LoS and allocating or deallocating nodes. Also, deciding whether a tuple is a sample tuple or not and inserting this tuple into the sample can be done with one simple function call. In total the interface of the algorithm consist only out of a few simple function calls that need be called from the existing code.
